
The work presented in this thesis was carried out in the period from September 2014 to July 2017, at the Department of Energy and Process Engineering at the Norwegian University of Science and Technology. The project has been funded through the SIMCOFLOW project, carried out by the SINTEF Materials and Chemistry. I gratefully acknowledge the financial support from the Research Council of Norway (project no.~234126/30). 

I wish to thank my supervisor, Professor Bernhard M\"{u}ller, for allowing and encouraging me to pursue my own ideas, while ensuring that I stay on planned schedule and agreed deadlines. I am especially thankful for a lot of patience during the beginning of my PhD when I had a lot to learn and catch up with. I would like to thank my co-supervisor Tore Fl{\aa}tten for all our scientific and nonscientific discussions. I am especially thankful for teaching me countless mathematical concepts and tools and for having endless faith in our work even when I didn't. I wish to thank my co-supervisor Marica Pelanti for arranging my stay at ENSTA ParisTech and for being a great host during my six months stay in Paris. My stay in Paris has been a valuable experience both professionally and personally. Further, Marica's knowledge on HLL(C) schemes has been an invaluable asset when I first started to work with this schemes. I also wish to thank my co-supervisor Ernst Arne Meese. We did not collaborate closely, but I appreciated our discussions and interest in my work.

These three years would have been much less fun without all my co-workers from the Department of Energy and Process Engineering. I am very thankful for all Green Room breaks, movie nights, dinners and skiing trips, but most of all, thank you for sharing with me the joys and excitement of doing sweet science. There are too many people to name them all, and I will point out {\O}yvind, Ehsan, Eskil and Anna who had a misfortune of sharing an office and an apartment with me. 

Lastly, I wish to thank all my friends back home in German for not forgetting me during my long absences, and I wish to thank Martina for her endless support in all my endeavors. 

\vspace*{2\baselineskip}

\hfill Trondheim, December 2018

\hfill Alexander Busch
